### 2015-05-15 Manuscript/Missing_mammals-draft.tex
Minor updates in the intro

%Key words: total evidence method, data structure, phylogenetic, living, fossil, topology
\bigskip

@@

\section{Introduction}
%We need fossils - dangerously paraphrased from the missing data paper
Studying both living and fossil taxa together in essential to fully understand macrovelutionary patterns and processes and is becoming increasingly common among evolutionary biologists \cite{jacksonwhat2006,quentaldiversity2010,dietlconservation2011,slaterunifying2013,fritzdiversity2013}. Combining the global biodiversity (living and fossil) in such studies allows for example, to improve accuracy of the timing of diversification events (e.g. \cite{ronquista2012}), to better understand relationships among lineages (e.g. \cite{beckancient2014}) or to infer biogeographical patterns through time (e.g. \cite{Meseguer01032015}).
%We can use TEM for that
In order to do so, one must efficiently combine living and fossil taxa data in macroevolutionary models. One trending method, called the total evidence method, allows to combine molecular data from living taxa and morphological data from both living and fossil taxa in a supermatrix (e.g. \cite{pyrondivergence2011,ronquista2012,schragocombining2013,slaterunifying2013,beckancient2014,Meseguer01032015}). This method allows to use all the available data and to treat fossil as tips rather than nodes and allows integrative phylogenetic inference methods such as tip-dating \cite{ronquista2012,Drummond01082012,BEASTmaster}.

%But missing data in living taxa problem
However, the total evidence method requires a lot of data, both molecular and morphological for both living and fossil taxa. Entire sections of this data can sometimes be difficult or impossible to collect for every taxa present in the analysis. For example, fossil have really rarely molecular data available and morphological characters are rarely collected when molecular data is available (e.g. \cite{O'Leary08022013} \textit{vs.} \cite{meredithimpacts2011}). Therefore, the ability of this method to recover correct topology is expected to decrease with missing data and especially when few morphological data overlaps between the living and the fossil taxa \cite{GuillermeCooper}. %Owwwww the sneaky self cite ;-).
The effect of missing data is most important on topology when too few living taxa have available morphological data \cite{GuillermeCooper}.
For example, if there is no morphological available for any living taxa of an entire clade, it is impossible to link a fossil taxon to this clade because no morphological data will overlap between the fossil (regardless the amount of data available for the fossil) and the living taxa (that have no morphological data). This property of the total evidence method can lead to important topological incongruities because the fossil will only be able to branch to a clade of living taxa that contains morphological data, even if in reality, the fossil taxon does not belong to that clade (see figure ~\ref{Figure_missing_data_problem}).

@@

%What we're doing
It is therefore crucial to understand the distribution of the available morphological data for living taxa in a clade before using a total evidence approach because missing living taxa with morphological data can lead to two topological artefacts: (1) the impossibility for a fossil to branch in the right clade if there is no morphological data available in this clade \cite{GuillermeCooper}; and (2) the higher probability for a fossil to branch within a wrong clade that has more morphological data available for living taxa than the right clade \cite{GuillermeCooper}.

In this study, we assess the level of available data in living mammals in order to highlight the two caveats described above.
%In this study we investigate the amount of living mammal taxa with available morphological data to detect the two potential caveats described above that could lead to topological errors when building total evidence mammal phylogenetic trees.
We extracted living mammals taxa with available morphological data from 256 phylogenetic matrices available online and measured the proportion of morphological data availability for each mammalian order. Additionally, in the mammalian orders where data was missing, we estimated the structure of the available data to detect if the available data was biases towards certain clades only or if it was randomly distributed using the Net Relatedness Index \cite{webb2002phylogenies}.

%Finaly using phylogenetic targeting (CITE), we proposed a strategy to fill the gaps.

@@

%\subsection{Data availability}
The amount of living taxa with no available morphological data was surprisingly high at the species level: 24/28 orders have less than 75\% taxa with available morphological data (and two of the 28 orders are monospecific). This threshold of 75\% available data is the maximum amount of possible missing data (25\%) before having a significant effect on the topology of total evidence trees \cite{GuillermeCooper}. Beyond this threshold, there is a noteworthy displacement of the wildcard taxa (\textit{sensu} \cite{kearneyfragmentary2002}) as well as a important decrease in the conservation of clades \cite{GuillermeCooper}. At the species level within most orders of mammals, it is therefore likely to see topological artefacts for the placement of fossil taxa. However, the data availability seems to be less an issue at higher taxonomic level (i.e. the Genus and the Family level). This point is important to consider from a practical point of view because of the slight discrepancy between neontological and palaeontological species. While neontological species are described on accurately measured morphology, genetic distance, spacial distribution and even behaviour (e.g. \cite{kellymolecular2014}); palaeontological data can be based only on morphological, spatial and temporal data (e.g. \cite{ni2013oldest}). Because of this difference between neontological species (\textit{sensu} reproductive isolates) and paleontological species (\textit{sensu} morpho-species), most palaeontological studies are using the Genus level for their smallest taxonomic OTU (e.g. \cite{ni2013oldest,O'Leary08022013}). In this frame of palaeontological data usage, the data availability at the Genus level in living mammals should be our priority concern.

@@

%Let's code some data
Following the recommendations in \cite{GuillermeCooper}, one should code morphological characters for a maximum of living species in order to improve the topology of total evidence trees containing both living and fossil taxa. Since the data for living mammals is usually easily available in world wide vast natural history collections, we propose that an increased effort in coding morphological characters from living species should be done via engaging in collaborative data collection projects through web portals such as \textit{morphobank} \cite{morphobank}.

>>>>

%Key words: Total Evidence method, data structure, phylogenetic, living, fossil, topology
\bigskip

@@

\section{Introduction}
%We need fossils - dangerously paraphrased from the missing data paper
Studying both living and fossil taxa together in essential to fully understand macrovelutionary patterns and processes and is becoming increasingly common among evolutionary biologists \cite{jacksonwhat2006,quentaldiversity2010,dietlconservation2011,slaterunifying2013,fritzdiversity2013,Wood01032013}. Combining the global biodiversity by including both living and fossil in such studies allows, for example, to improve accuracy of the timing of diversification events (e.g. \cite{ronquista2012}), to better understand relationships among lineages (e.g. \cite{beckancient2014}) or to infer biogeographical patterns through time (e.g. \cite{Meseguer01032015}). %And add STD paper? Or that's too much self cites?
%We can use TEM for that
In order to perform such analysis, one must efficiently combine living and fossil taxa data in macroevolutionary models. One trending method, called the Total Evidence method \cite{eernissetaxonomic1993,ronquista2012}, allows to combine molecular data from living taxa and morphological data from both living and fossil taxa in a supermatrix (e.g. \cite{pyrondivergence2011,ronquista2012,schragocombining2013,slaterunifying2013,beckancient2014,Meseguer01032015}). This method not only allows to use all the available data but also allows to treat fossil taxa as tips rather than nodes \textit{via} integrative phylogenetic inference methods such as tip-dating \cite{ronquista2012,Drummond01082012,Wood01032013,BEASTmaster}.

%But missing data in living taxa problem
However, the Total Evidence method requires, by definition, a lot of data. One must collect both molecular data for living taxa as well as morphological data for both living and fossil taxa, two types of data that require fairly different technical skills (e.g. \cite{meredithimpacts2011} \textit{vs.} \cite{O'Leary08022013}). Additionally, entire sections of this data can sometimes be difficult or impossible to collect for every taxa present in the analysis. For example, fossil have really rarely molecular data available and morphological characters are rarely collected when molecular data is available. Therefore, the ability of the Total Evidence method to recover correct topology is expected to decrease when there is a low overlap between the living and the fossil data \cite{GuillermeCooper}. The effect of missing data is most important on topology when too few living taxa have available morphological data \cite{GuillermeCooper}. For example, if there is no morphological available for any living taxa within an entire clade, it is impossible to link a fossil taxon to this clade because no morphological data will overlap between the fossil (regardless the amount of data available for the fossil) and the living taxa (that have no morphological data). This property of the Total Evidence method can rapidly lead to important topological incongruities because the fossil will only be able to branch to a clade of living taxa that contains morphological data, even if in reality, the fossil taxon does not belong to that clade \cite{GuillermeCooper} (see figure ~\ref{Figure_missing_data_problem}).

@@

%What we're doing
It is therefore crucial to understand the distribution of the available morphological data for living taxa in a clade before using a Total Evidence approach because missing living taxa with morphological data can lead to two topological artefacts: (1) the impossibility for a fossil to branch in the right clade if there is no morphological data available in this clade \cite{GuillermeCooper}; and (2) the higher probability for a fossil to branch within a wrong clade that has more morphological data available for living taxa than the right clade \cite{GuillermeCooper}. In this study, we assess the level of available morphological data in living mammals in order to highlight the two caveats described above.
%In this study we investigate the amount of living mammal taxa with available morphological data to detect the two potential caveats described above that could lead to topological errors when building Total Evidence mammal phylogenetic trees.
We collected living mammals morphological data from @256 phylogenetic matrices available online and measured the proportion of morphological data availability for each mammalian order. Additionally, in the mammalian orders where data was missing, we estimated the structure of the available data to detect if the available data was biased towards certain clades only or if it was randomly distributed using the Net Relatedness Index \cite{webb2002phylogenies}.

@@

%\subsection{Data availability}
The amount of living taxa with no available morphological data was surprisingly high at the species level: 24/28 orders have less than 75\% taxa with available morphological data (and two of the 28 orders are monospecific). This threshold of 75\% available data is the maximum amount of possible missing data (25\%) before having a significant effect on the topology of Total Evidence trees \cite{GuillermeCooper}. Beyond this threshold, there is a noteworthy displacement of the wildcard taxa (\textit{sensu} \cite{kearneyfragmentary2002}) as well as a important decrease in the conservation of clades \cite{GuillermeCooper}. At the species level within most orders of mammals, it is therefore likely to see topological artefacts for the placement of fossil taxa. However, the data availability seems to be less an issue at higher taxonomic level (i.e. the Genus and the Family level). This point is important to consider from a practical point of view because of the slight discrepancy between neontological and palaeontological species. While neontological species are described on accurately measured morphology, genetic distance, spacial distribution and even behaviour (e.g. \cite{kellymolecular2014}); palaeontological data can be based only on morphological, spatial and temporal data (e.g. \cite{ni2013oldest}). Because of this difference between neontological species (\textit{sensu} reproductive isolates) and paleontological species (\textit{sensu} morpho-species), most palaeontological studies are using the Genus level for their smallest taxonomic OTU (e.g. \cite{ni2013oldest,O'Leary08022013}). In this frame of palaeontological data usage, the data availability at the Genus level in living mammals should be our priority concern.

@@

%Let's code some data
Following the recommendations in \cite{GuillermeCooper}, one should code morphological characters for a maximum of living species in order to improve the topology of Total Evidence trees containing both living and fossil taxa. Since the data for living mammals is usually easily available in world wide vast natural history collections, we propose that an increased effort in coding morphological characters from living species should be done via engaging in collaborative data collection projects through web portals such as \textit{morphobank} \cite{morphobank}.
